Program Equivalence : Relational Separation Logic interactive prover implemented in Maude

### 2018-06-10 paper/paper.tex
finished hoare + sep logic paper parts'

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{array}
\usepackage{tabularx}
\author{ANDREI-ALIN CORODESCU}
\title{Program Equivalence : Relational Separation Logic interactive prover implemented in Maude}
\begin{document}
\begin{titlepage}
\begin{center}
\textbf{
"ALEXANDRU IOAN CUZA" UNIVERSITY OF IAȘI
}
\\
\textbf{FACULTY OF COMPUTER SCIENCE}
\end{center}
   \vspace{40mm}
\begin{center}
	\Large\textbf {Program Equivalence : Relational Separation Logic interactive prover implemented in Maude}\\
	\vspace{40mm}
	\large\textit {ANDREI-ALIN CORODESCU}
	\\
	\vspace{20mm}
	\textbf{Session: }\textit{July, 2018}\\
	\vspace{30mm}
	\textbf{Scientific Coordinator}\\
	\textbf{\textit{Conf. Dr. Ciobâcă Ștefan}}
	\vspace{30mm}
\end{center}
\end{titlepage}
\tableofcontents
\pagebreak

@@

A \(Heap\) represents a mapping from the \(\mathtt{PosNats}\) to \(\mathtt{Integers}\)
\[ \mathtt{H} : \mathtt{PosNats} \rightarrow \mathtt{Integers} \] 
More informally, the \(Store\) holds the value of the variables while the \(Heap\) maps the active memory cells during a program execution to their contents.
\subsubsection{Syntax and Semantics of the language}
\begin{figure}[h]
	\begin{tabularx}{\linewidth}{l  X}
	Integer Expressions& \(E ::= x \mid Integer \mid E\ \mathtt{plus}\ E \mid E\ \mathtt{times}\ E \mid E\ \mathtt{minus}\ E \) \\
	\\
	Boolean Expressions& \(B ::= false \mid true \mid B\ \mathtt{\&\&}\ B \mid  B\ ||\ B \mid B\  \mathtt{->}\ B \mid B \mathtt{<=>} B \mid \ ! B \mid E\ \mathtt{ge}\ E \mid E\ \mathtt{le}\ E \mid E\ \mathtt{eqs}\ E \)\\
	\\ 	
	Commands& \(C ::= x := alloc(E) \mid x := [E] \mid [E] := E \mid free(E) \mid x := E \mid C;C \mid if\ B\ then\ C\ else\ C \mid while\ B\ do\ C\ od \) \\
	\end{tabularx}
	\caption{Syntax of the language}
	\label{fig:langSyntax}
\end{figure}
The syntax of the language is presented in Figure \ref{fig:langSyntax} . It represents an adapted subset of the language presented in the Relational Separation Logic paper \cite{relational} . 
\paragraph{Semantics of the language}
The semantics of the previously defined language are standard but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:
\begin{itemize}
	\item{Verbosity of operators - most operators are replaced by their literal names ( \(+\) becomes \(\mathtt{plus}\) etc.); we have opted for this approach to avoid complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on Maude's primitives.}
		\subitem{\( B\ \mathtt{ge}\ B\) translates to the usual \(>=\) operator}
		\subitem{\( B\ \mathtt{le}\ B\) translates to the usual \(<=\) operator}
		\subitem{\( B\ \mathtt{eqs}\ B\) translates to the usual \(==\) equality comparison operator}
		\subitem{\( B\ \mathtt{->}\ B\) denotes implication between boolean expressions}
		\subitem{\(B\ \mathtt{<=>}\ B\) denotes equivalence between boolean expressions}
		\subitem{The rest of the operators are mostly similar to their C++ or Java counterparts and their semantics are self explanatory}
	\item{Memory allocation / deallocation: }
		\subitem{\(x := alloc(E)\) allocates a new cell in the memory, initializes it with the value of \(E\) and stores its address in the variable \(x\)}
		\subitem{\(x := free(E)\) deallocates the cell at the address equal to the value of \(E\)}
	\item{Working with variables and memory: }
		\subitem{\( x := [E] \) reads the contents of the memory cell at address \(E\) and stores the value in the variable \(x\)}
		\subitem{\( [E] := E\prime \) updates the contents of the memory cell at address \(E\) with the value of the expression \(E\prime\)}
		\subitem{\( x := E \) updates the value of the variable \(x\) with the value of the expression \(E\) - note that this command does not modify the heap in any way, as it usually happens with regular programming languages when updating the value of a variable}
\end{itemize}
\subsection{Hoare Logic}
Hoare Logic is a formal system which allows us to rigorously reason about the behaviour of programs.
The main component of Hoare Logic is the Hoare Triple, defined as follows:
\[ \{P\}\  C\  \{Q\} \]
where \(P\) denotes the state of the program before the execution of the command \(C\) and \(Q\) denotes the state of the program after the execution of \(C\). \\

@@

\(P\) and \(Q\), called precondition and postcondition, respectively can be specified using an assertion language . Figure \ref{fig:assertionSyntax} represents the assertion language, defined in relation with the language syntax: 
\vspace{1cm}
\begin{figure}
	\begin{tabularx}{\linewidth}{l  X}
		Assertions& \(P ::= B \mid P\ \Rightarrow\ P \mid P \land P \mid P \lor P \mid \neg P \mid {\exists x . P} \) \\
	\end{tabularx}
	\caption{Syntax of the assertion language}
	\label{fig:assertionSyntax}
\end{figure}
\\

At the base of proofs based on Hoare Logic are a set of axioms and proof rules which can be applied to Hoare triples. These axioms and rules will be discussed in the following section.
\subsection{Separation Logic}
Separation Logic is an extension of Hoare Logic which simplifies reasoning about program execution by introducing ways of localising parts of the proof : draw conclusions based on a subset of the program state, the subset chosen such that it contains only the parts relevant to the command specified by the analysed Hoare triple. \\

@@

\begin{figure}
	\begin{tabularx}{\linewidth}{l  X}
		Assertions& \(P ::= ... (specified\ previously) \mid E \mapsto E \mid P * P \mid emp \) \\
	\end{tabularx}
	\caption{Syntax of the assertion language}
	\label{fig:SLassertionSyntax}
\end{figure}

\paragraph{Semantics of the new elements}
\begin{itemize}
	\item{\(E \mapsto E\prime \) means that the heap has an active cell at address \(E\), with the content \(E\prime\), and the cell is the \textbf{only} active cell in the heap}
	\item{\(P * Q\) holds for heap \(h\) iff \(\exists h_0,h_1\) such that \(h_0 * h_1 = h\) and \(P\) holds for \(h_0\) and \(Q\) holds for \(h_1\)}
	\item{\(emp\) denotes a heap without any active cells}
\end{itemize}
\subsection{Relational Separation Logic}
\subsection{Rewriting Logic}
\section{Relational Separation Logic Interactive Prover}
Ordinea subsectiunilor mai poate fi schimbata ulterior. Capitolul cel mai semnificativ din lucrare, va contine toate detaliile de implementare ale solutiei curente
\subsection{Maude - General Overview}
\subsection{Executable semantics of the language}
\subsection{Modelling the Separation Logics}
\subsection{Interaction by Maude LOOP-MAUDE}
\subsection{Prover execution flow}
\subsection{Automated processes}
\subsubsection{Automatic matching of axioms and previously proven goals}
\subsubsection{Automatic demonstration of implications}
\subsection{User Interface}
\subsection{Future directions}
\section{Additional Research}
\subsection{Automatic prover}
\subsection{Concurrent programs extension}
\subsection{Java+ITP}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{array}
\usepackage{tabularx}
\usepackage{ebproof}
\author{ANDREI-ALIN CORODESCU}
\title{Program Equivalence : Relational Separation Logic interactive prover implemented in Maude}
\begin{document}
\begin{titlepage}
\begin{center}
\textbf{
"ALEXANDRU IOAN CUZA" UNIVERSITY OF IAȘI
}
\\
\textbf{FACULTY OF COMPUTER SCIENCE}
\end{center}
   \vspace{40mm}
\begin{center}
	\Large\textbf {Program Equivalence : Relational Separation Logic interactive prover implemented in Maude}\\
	\vspace{40mm}
	\large\textit {ANDREI-ALIN CORODESCU}
	\\
	\vspace{20mm}
	\textbf{Session: }\textit{July, 2018}\\
	\vspace{30mm}
	\textbf{Scientific Coordinator}\\
	\textbf{\textit{Conf. Dr. Ciobâcă Ștefan}}
	\vspace{30mm}
\end{center}
\end{titlepage}
\tableofcontents
\pagebreak

@@

A \(Heap\) represents a mapping from the \(\mathtt{PosNats}\) to \(\mathtt{Integers}\)
\[ \mathtt{H} : \mathtt{PosNats} \rightarrow \mathtt{Integers} \] 
More informally, the \(Store\) holds the value of the variables while the \(Heap\) maps the active memory cells during a program execution to their contents.
\subsubsection{Syntax and Semantics of the language}
\begin{figure}[h]
	\noindent\rule{\linewidth}{0.4pt}
	\begin{tabularx}{\linewidth}{l  X}
	Integer Expressions& \(E ::= x \mid Integer \mid E\ \mathtt{plus}\ E \mid E\ \mathtt{times}\ E \mid E\ \mathtt{minus}\ E \) \\
	\\
	Boolean Expressions& \(B ::= false \mid true \mid B\ \mathtt{\&\&}\ B \mid  B\ ||\ B \mid B\  \mathtt{->}\ B \mid B \mathtt{<=>} B \mid \ ! B \mid E\ \mathtt{ge}\ E \mid E\ \mathtt{le}\ E \mid E\ \mathtt{eqs}\ E \)\\
	\\ 	
	Commands& \(C ::= x := alloc(E) \mid x := [E] \mid [E] := E \mid free(E) \mid x := E \mid C;C \mid if\ B\ then\ C\ else\ C \mid while\ B\ do\ C\ od \) \\
	\end{tabularx}
	\caption{Syntax of the language}
	\label{fig:langSyntax}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}
The syntax of the language is presented in Figure \ref{fig:langSyntax} . It represents an adapted subset of the language presented in the Relational Separation Logic paper \cite{relational} . 
\paragraph{Semantics of the language}
{
The semantics of the previously defined language are standard but a few clarifications are necessary to point out how our language constructs interact with the storage model, as well as a few differences from regular programming languages:
\begin{itemize}
	\item{Verbosity of operators - most operators are replaced by their literal names ( \(+\) becomes \(\mathtt{plus}\) etc.); we have opted for this approach to avoid complication when implementing the language in Maude, because most of the operators are already defined in Maude for its built-in types, and overloading them could cause conflicts since we are basing our own defined types on Maude's primitives.}
		\subitem{\( B\ \mathtt{ge}\ B\) translates to the usual \(>=\) operator}
		\subitem{\( B\ \mathtt{le}\ B\) translates to the usual \(<=\) operator}
		\subitem{\( B\ \mathtt{eqs}\ B\) translates to the usual \(==\) equality comparison operator}
		\subitem{\( B\ \mathtt{->}\ B\) denotes implication between boolean expressions}
		\subitem{\(B\ \mathtt{<=>}\ B\) denotes equivalence between boolean expressions}
		\subitem{The rest of the operators are mostly similar to their C++ or Java counterparts and their semantics are self explanatory}
	\item{Memory allocation / deallocation: }
		\subitem{\(x := alloc(E)\) allocates a new cell in the memory, initializes it with the value of \(E\) and stores its address in the variable \(x\)}
		\subitem{\(x := free(E)\) deallocates the cell at the address equal to the value of \(E\)}
	\item{Working with variables and memory: }
		\subitem{\( x := [E] \) reads the contents of the memory cell at address \(E\) and stores the value in the variable \(x\)}
		\subitem{\( [E] := E\prime \) updates the contents of the memory cell at address \(E\) with the value of the expression \(E\prime\)}
		\subitem{\( x := E \) updates the value of the variable \(x\) with the value of the expression \(E\) - note that this command does not modify the heap in any way, as it usually happens with regular programming languages when updating the value of a variable}
\end{itemize}
}
\subsection{Hoare Logic}
Hoare Logic is a formal system which allows us to rigorously reason about the behaviour of programs.
The main component of Hoare Logic is the Hoare Triple, defined as follows:
\[ \{P\}\  C\  \{Q\} \]
where \(P\) denotes the state of the program before the execution of the command \(C\) and \(Q\) denotes the state of the program after the execution of \(C\). \\

@@

\(P\) and \(Q\), called precondition and postcondition, respectively can be specified using an assertion language . Figure \ref{fig:assertionSyntax} represents the assertion language, defined in relation with the language syntax: 
\vspace{1cm}
\begin{figure}[h]
	\noindent\rule{\linewidth}{0.4pt}
	\begin{tabularx}{\linewidth}{l  X}
		Assertions& \(P ::= B \mid P\ \Rightarrow\ P \mid P \land P \mid P \lor P \mid \neg P \mid {\exists x . P} \) \\
	\end{tabularx}
	\caption{Syntax of the assertion language}
	\noindent\rule{\linewidth}{0.4pt}
	\label{fig:assertionSyntax}
\end{figure}
\\

\begin{figure}[h!]
	\noindent\rule{\linewidth}{0.4pt}
	Empty Statement Axiom
	\[
	\begin{prooftree}
	\Hypo{}
	\Infer 1 {\{P\}\  \mathtt{skip}\  \{P\}}
	\end{prooftree}	
	\]
	Assignment Axiom 
	\medskip
	\[
	\begin{prooftree}
	\Hypo{}
	\Infer 1 {\{P[E/x]\}\  \mathtt{x := E}\  \{P\}}
	\end{prooftree}	
	\]
	Composition Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{ \{P\} S \{Q\}}
	\Hypo{ \{Q\} T \{R\}}
	\Infer 2 {\{P\}\  \mathtt{S ; T}\  \{Q\}}
	\end{prooftree}	
	\]
	Consequence Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q \Rightarrow Q\prime}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Auxiliary Variable Elimination Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{\{P\} C \{Q\}}
	\Infer 1 [\(x \notin Free(C)\)]{\{\exists x.P\}\ C\ \{\exists x. Q\}}
	\end{prooftree}	
	\]
	Conditional Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{\{B \land P\} S \{Q\}}
	\Hypo{\{\neg B \land P\} T \{Q\}}
	\Infer 2 {\{P\} \mathtt{if\ B\ then\ S\ else\ T\ fi} \{Q\}}
	\end{prooftree}	
	\]
	While Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{\{P \land B\} S\{P\}}
	\Infer 1 {\{P\}\  \mathtt{while\ B\ do\ S\ od}\  \{\neg B \land P\}}
	\end{prooftree}	
	\]
	\caption{Hoare Logic Rules}
	\label{fig:HLRules}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}

At the base of proofs based on Hoare Logic are a set of axioms and proof rules which can be applied to Hoare triples, presented in Figure \ref{fig:HLRules}

\subsection{Separation Logic}
Separation Logic is an extension of Hoare Logic which simplifies reasoning about program execution by introducing ways of localising parts of the proof : draw conclusions based on a subset of the program state, the subset chosen such that it contains only the parts relevant to the command specified in the analysed Hoare triple. \\

@@

\begin{figure}
	\noindent\rule{\linewidth}{0.4pt}
	\begin{tabularx}{\linewidth}{l  X}
		Assertions& \(P ::= ... (specified\ previously) \mid E \mapsto E \mid P * P \mid emp \) \\
	\end{tabularx}
	\caption{Syntax of the assertion language}
	\label{fig:SLassertionSyntax}
	\noindent\rule{\linewidth}{0.4pt}
\end{figure}

\paragraph{Semantics of the new elements}
\begin{itemize}
	\item{\(E \mapsto E\prime \) means that the heap has an active cell at address \(E\), with the content \(E\prime\), and the cell is the \textbf{only} active cell in the heap}
	\item{\(P * Q\) holds for heap \(h\) iff \(\exists h_0,h_1\) such that \(h_0 * h_1 = h\) and \(P\) holds for \(h_0\) and \(Q\) holds for \(h_1\)}
	\item{\(emp\) denotes a heap without any active cells}
\end{itemize}
\paragraph{Inference Rules for Separation Logic} 
Separation Logic introduces a new rule, called Frame Rule, which makes use of the \(*\) operator to allow local reasoning :
\\
\noindent\rule{\linewidth}{0.4pt}
Frame Rule
\medskip
\[
\begin{prooftree}
\Hypo{\{P\}\ C\ \{Q\}}
\Infer 1 [\(Modifies(C) \cup Free(R) = \emptyset \)]{\{P * R \}\ C\ \{Q * R\}}
\end{prooftree}		
\]
\noindent\rule{\linewidth}{0.4pt}
\subsection{Relational Separation Logic}
Relational Separation Logic builds upon the foundations layed by the Separation Logic by extending it to facilitate analysing how two programs are related. 
\paragraph{Hoare Quadruple}
The main abstraction introduced by the Relational Separation Logic is the Hoare Quadruple, which express the effects of commands on the states of the two programs. 
\[\{R\}\ {C_1 \atop C_2}\   \{T\}\]

\(R\) represents a \textbf{relation} between the two program states holding before the execution of \(C_1\) and \(C_2\) respectively while \(T\) represents a \textbf{relation} which holds after the execution of the two commands. 
\subsection{Rewriting Logic}
\section{Relational Separation Logic Interactive Prover}
Ordinea subsectiunilor mai poate fi schimbata ulterior. Capitolul cel mai semnificativ din lucrare, va contine toate detaliile de implementare ale solutiei curente
\subsection{Maude - General Overview}
\subsection{Executable semantics of the language}
\subsection{Modelling the Separation Logics}
\subsection{Interaction by Maude LOOP-MAUDE}
\subsection{Prover execution flow}
\subsection{Automated processes}
\subsubsection{Automatic matching of axioms and previously proven goals}
\subsubsection{Automatic demonstration of implications}
\subsection{User Interface}
\subsection{Future directions}
\section{Additional Research}
\subsection{Automatic prover}
\subsection{Concurrent programs extension}
\subsection{Java+ITP}
